A Dearth of Young Stars and Planets in the Kepler Field

### 2024-05-23 paper/ms.tex
update ms

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.475\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{{\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (gray points).  The
    rotation periods are primarily drawn from
    \citet{Santos_2019,Santos_2021}.  About half of these rotators are
    suitable for gyrochronology (dark gray points), based on factors
    including their non-binarity and
    proximity to the spectroscopic main sequence (orange boundary; see Section~\ref{subsec:flags}).
    Some are ``confirmed'' or ``candidate'' KOIs
    that meet an additional set of planetary quality criteria (yellow
    points; Section~\ref{subsec:plflags}).  Surface gravities and
    effective temperatures were derived photometrically by
    \citet{Berger_2020a_catalog}.
    Isochrones 
    in the lower panel are from MIST \citep{Choi_2016} .
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.25cm}
	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}.  The
    intrinsic dispersion becomes much larger
    than changes in the mean at $\gtrsim$1\,Gyr.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added concerns of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the exact details of how this locus is constructed are arbitrary\footnote{For purposes of reproducibility, 
	the method was as follows.
The orange line in Figure~\ref{fig:stellarprops} is	defined as
the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
We defined $f_0$ by fitting an
$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
and $T_{\rm eff}\in[3700,5900]$\,K, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  
By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
However for F and G dwarfs, the resulting locus
allowed for stars that were `too evolved''; the density of KOIs 
was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
We therefore visually selected KOIs that appeared
to be near the main sequence -- i.e.~ most of the yellow
points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
polynomial $f_1$ through a similar BIC-minimization procedure,
which yielded $N$=3.
This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
The portion of the orange locus from $\approx$5300--6200\,K
is set as $f_1 + c$, for $c$ a constant offset that we visually tuned to be
0.1\,dex above the ``KOI main sequence''.
The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
We emphasize that we do report rotation-based ages for stars
above the orange locus;
they are simply flagged as not being near the main sequence.  }
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

The point of Figure~\ref{fig:agescalecompone} is hopefully clear.
While previous studies have been able to derive accurate ages for
stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have
historically been overestimated by anywhere from 0.3--2\,dex.  The
situation is particularly severe for the 40\,Myr stars in Cep-Her, but
it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
For isochrone and kinematic ages, this is not surprising because these
methods rely on parameters that do not appreciably change at
$t$$\lesssim$1\,Gyr.  For the rotation-based ages quoted by
\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
discrepancy is likely
caused by details of the calibration methodologies.  The physical
spin-down models used by \citet{2023ApJ...952..131M}, and the
gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
may not reproduce the rotation sequences of open clusters such as
Praesepe, the Hyades, and the Pleiades.  The \citet{Mamajek_2008}
calibration used by \citet{Reinhold_2015} does reproduce such
sequences for Sun-like stars.  However the
uncertainties determined using this approach are underestimates, because
they do not marginalize over
the intrinsic scatter in rotation periods present for stars of a fixed
mass and age.

@@

Let us restrict our attention to blemish-free stars, with blemish-free
``confirmed'' planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must have no quality flags
raised at all.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.

These planets are shown in Figure~\ref{fig:rp_period_age_results}.
They include \nearthshighqconfirmed\ Earth-sized or smaller planets,
no Jupiter-sized or larger planets, \nsubsaturnshighqconfirmed\
``super-Neptune'' ($>$4\,$R_\oplus$) sized planets,
\nminineptuneshighqconfirmed\ mini-Neptunes above the radius valley
(as parametrized by \citealt{2018MNRAS.479.4786V}), and
\nsuperearthshighqconfirmed\ super-Earth sized planets below the
radius valley.  While most of these planets are on orbital periods
below 50\,days, \nlongperiodhighqconfirmed\ are on more distant
orbits.

A few new highlights from the high quality stars and planets include
Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation
and lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
near-Solar analog with two transiting planets: an Earth-sized planet
on a one-day orbit, and a mini-Neptune on a five-day orbit.  The
Earth-sized planet, Kepler-1312c, along with the near-identical
Kepler-1561 b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
youngest Earth-sized planets currently known, comparable to systems
such as HD~63433d \citep[1.1\,$R_\oplus$,
$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b
\citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
Kepler-1629~b, with a size just two-thirds that of Earth
(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on
rotation (\kepsixteentwonine\,Myr).  The lithium age from a
reconnaissance TRES spectrum confirms the point.

@@

Most of the stars observed by Kepler are in the thin disk.  This can
be verified by following \citet{Gaia_2018}, and denoting stars with 2D
tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
and those with 60$<$$v_{\rm T}$$<$150\,\kms as thick disk members.
For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
disk stars respectively.  The thin disk stars have a threefold larger
detected rotation period fraction: 35{,}675 rotators are in the thin
disk, while 7{,}312 are in the thick disk.  Further imposing the
gyrochronology quality flags discussed in Section~\ref{subsec:flags}
yields 17{,}755 thin-disk stars for which gyrochronology is nominally
applicable, and 2{,}462 thick-disk stars.

@@

% compare against Berger+20, 
% note their sentence
% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al. (2018) and Pinsonneault et al. (2018), as well as the rotation-based ages in Claytor et al. (2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al. (2019)."
%
% also compare against Mor2019 and their literature assessment
% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al. (2015), and Haywood et al. (2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation. Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly. The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al. (2002) and Cignoni et al. (2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively. Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen_Lai_2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.
Some care is required since beyond age,
exoplanet demographics also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018,Miyazaki2023}.

@@

We began this article with two questions: how wrong is the assumption
of a uniform age distribution for stars in the galactic thin disk?
And why are only $\approx$50 sub-gigayear transiting planets known,
rather than the $\approx$500 that would be expected under the
assumption of a uniform star formation history for the
$\approx$5{,}000 known planets?

@@

Our analysis blindly recovered the youth of all 14 known Kepler planets in
clusters, with ages that agreed with the more precise ensemble ages.
While lithium provided only marginal added information for
most of the sample, in cases for which comparison was possible, the independent lithium and rotation ages agreed
$\gtrsim$90\% of the time.  In particular, we reported {\bf XX}
confirmed $<$500\,Myr planets with strong age constraints from both
rotation and lithium, including {\bf YY} mini-Neptunes, {\bf ZZ}
super-Earths, {\bf AA} Earths and sub-Earths, and a Neptune-sized
planet.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item The age distributions of both the Kepler target stars and the
    known Kepler planet-hosts show a demographic cliff.  There are
    \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr) stars in the
    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
    three billion years ago.
    This star formation history broadly agrees with those derived
    from CMD fitting and white-dwarf chronology, though with
    important nuances.
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the
    Kepler field implies that we should instead expect $\approx$250.
  %
  %FIXME: 120 number
  \item Using rotation-based ages, we have derived ages for
    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
    and \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$120-170 known sub-gigayear planets with well-constrained
    ages.  This lessens the original factor of ten discrepancy to
    a factor of at most two.
\end{enumerate}

@@

\appendix
\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
\label{app:inconsistent}

@@

\paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} has
a G0 V primary and a $\approx$2.5\,day near-grazing transiting hot
Jupiter, and has been studied in detail by multiple investigators
\citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
Our lithium-based age of this system (\trestwotli\,Myr), derive from
EW$_{\rm Li}$=81$\pm$3\,m\AA, is qualitatively in line with the
previously noted lithium abundance \citep{2007ApJ...664.1190S}.
However, we detect no photometric rotation signal, and the
spectroscopic $v\sin i$ is low.  The lithium age is strongly
asymmetric because there is significant amounts of scatter in EW$_{\rm
Li}$ vs.~$T_{\rm eff}$ over ages $t$$\gtrsim$1\,Gyr for early G
dwarfs.  We calculate a $2\sigma$ upper limit on $t_{\rm Li}$ of
2.8\,Gyr, and a $3\sigma$ upper limit of 6.1\,Gyr.  The non-detection
of rotation, while mildly surprising, is not shocking if the system's
true age is in this long tail.

\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm
Li}$=100$\pm$5\,m\AA, nominally implying $t_{\rm
Li}$$\approx$200-800\,Myr, but no rotation detection.  Our automated
flags noted that this star has a low surface gravity, high luminosity,
and is far from the main sequence.  \citet{2014A&A...572A...2B}
concur: this star is a sub-giant slightly more massive than the Sun;
no rotation detection is expected; the lithium likely simply survived
over the star's main sequence lifetime.

@@

\paragraph{Kepler-786}---This K3 V star has a surprisingly high
lithium content.  The spectrum is single-lined, with
\texttt{SpecMatch-Synth} derived parameters of $T_{\rm
eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm Li}$=83$\pm$6\,m\AA, the
lithium age of \kepseveneightsix\,Myr would predict an obvious
rotation signal.  None is present in the Kepler light curve,
consistent with the low $v\sin i$.  Out of all ``apparently
discrepant'' lithium and rotation measurements discussed in this
appendix, this is the only one that seems to remain discrepant after
scrutiny.  The  \ion{Ca}{2} doublet is in emission, with $R'_{\rm
HK}=-4.7\pm0.5$, which suggests an age at least as old as the Hyades
\citep{Mamajek_2008}.  The Balmer lines are in absorption, and display
no obvious signatures of youth.

@@

\section{TODO's}

\begin{enumerate}
%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
%    Could compare against asteroseis too.
%    %
%  \item Intro needs better discussion of Mathur2023 and Lu2021
    %
  \item Consider moving the bit list description to its own appendix.
    %
  \item Rerun gyro calculation with denser grid, fixing quantization
    and extending upper limit to 5gyr.  (Will use updated Teff
    uncertainties)
\end{enumerate}

>>>>

		\vspace{-0.3cm}
		\subfloat{
			\includegraphics[width=0.475\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
  \caption{{\bf The stars.}  Our analysis focuses on stars observed by
  Kepler with previously reported rotation periods (gray points).  The
  rotation periods are primarily drawn from
  \citet{Santos_2019,Santos_2021}.  About half of these rotators are
  suitable for gyrochronology (dark gray points), based on factors
  including their non-binarity and proximity to the spectroscopic main
  sequence (orange boundary; see Section~\ref{subsec:flags}).  Some
  are ``confirmed'' or ``candidate'' KOIs that meet an additional set
  of planetary quality criteria (yellow points;
  Section~\ref{subsec:plflags}).  Surface gravities and effective
  temperatures were derived photometrically by
  \citet{Berger_2020a_catalog}.  Isochrones in the lower panel are
  from MIST \citep{Choi_2016}.
	}
	\label{fig:stellarprops}
\end{figure}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.25cm}
	\caption{{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
    for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected from 2009--2024.  Lines are the ``mean''
    isochrones from \citet{Jeffries_2023}.  The
    intrinsic dispersion around these isochrones becomes much larger
    than changes in the mean at $\gtrsim$1\,Gyr (see
    Figure~\ref{fig:models}).  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

{\it Near the main sequence (bit 9)}---Many stars observed by Kepler
are far from the main sequence: this can be seen in both panels of
Figure~\ref{fig:stellarprops}.  This is a challenge for rotation-based
ages because {\it i)} point sources far from the main sequence include
unresolved photometric binaries, and {\it ii}) effects other than
magnetic braking influence the rotation periods of bona fide single
stars far from the main sequence.  The added concerns of
reddening and metallicity further complicate matters.  After exploring
various options, we settled on the orange locus in the $\log
g$--$T_{\rm eff}$ plane shown in Figure~\ref{fig:stellarprops}.  While
the exact details of how this locus is constructed are arbitrary (see Appendix~\ref{app:linemethod}),
the general aim is to flag stars for which there is evidence based on
their location in the $\log g$--$T_{\rm eff}$ plane that our
gyrochronology model may not be reliable.  
%The rotation period
%distribution of sources excluded in this manner includes a relatively
%larger fraction of ``rapid rotators'' than the main sequence sources,
%likely due to a combination of astrophysical and systematic effects.
See for instance Figure~8a of \citet{2022AJ....164..137K} or
Figures~12 and 13 of \citet{2023ApJS..268....4F} for examples of how
the period distribution of this population differs from stars near the
main sequence.

@@

The point of Figure~\ref{fig:agescalecompone} is hopefully clear.
While previous studies have been able to derive accurate ages for
stars $\gtrsim$1\,Gyr old, the ages of sub-gigayear stars have
historically been overestimated by anywhere from 0.3--2\,dex.  The
situation is particularly severe for the 40\,Myr stars in Cep-Her, but
it is also present for 150--300\,Myr stars in Melange-3 and Theia~520.
For isochrone and kinematic ages, this is not surprising because these
methods rely on parameters that do not appreciably change at
$t$$\lesssim$1\,Gyr.  For the rotation-based ages quoted by
\citet{2023ApJ...952..131M} and \citet{2024AJ....167..159L}, the
discrepancy is likely
caused by details of the calibration methodologies.  The physical
spin-down models used by \citet{2023ApJ...952..131M}, and the
gyro-kinematic calibration ages used by \citet{2024AJ....167..159L},
may not reproduce the rotation sequences of open clusters such as
Praesepe, the Hyades, and the Pleiades.  The \citet{Mamajek_2008}
calibration used by \citet{Reinhold_2015} does reproduce such
sequences for Sun-like stars.  However the
uncertainties determined using that method are underestimates because
they do not marginalize over
the intrinsic scatter in rotation periods present for stars of a fixed
mass and age.

@@

Let us restrict discussion to blemish-free stars with 
confirmed planets: $Q_{\rm star}$ must not have any of bits zero
through nine raised, and $Q_{\rm planet}$ must also have no quality flags
raised.  Under this constraint, our results include
\ltonegyrhighqconfirmedtwosided\ systems younger than 1\,Gyr with
two-sided $t_{\rm gyro}$ and $t_{\rm Li}$, and
\ltonegyrhighqconfirmedonesided\ cases for which $t_{\rm gyro}$
provides an age constraint while $t_{\rm Li}$ is one-sided.

These planets are shown in Figure~\ref{fig:rp_period_age_results}.
They include \nearthshighqconfirmed\ Earth-sized or smaller planets,
no Jupiter-sized or larger planets, \nsubsaturnshighqconfirmed\
``super-Neptune'' ($>$4\,$R_\oplus$) sized planets,
\nminineptuneshighqconfirmed\ mini-Neptunes above the radius valley
as parametrized by \citealt{2018MNRAS.479.4786V}, and
\nsuperearthshighqconfirmed\ super-Earth sized planets below the
radius valley.  While most of these planets are on orbital periods
below 50\,days, \nlongperiodhighqconfirmed\ are on more distant
orbits.

A few highlights from these systems include
Kepler-1529, Kepler-1565, Kepler-1312, and Kepler-1629.  Kepler-1529b
is a $\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation
and lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
near-Solar analog with an Earth-sized planet
on a one-day orbit, and a mini-Neptune on a five-day orbit.  The
Earth-sized planet, Kepler-1312~c, along with the near-identical
Kepler-1561~b ($t_{\rm gyro}$=\kepfifteensixone\,Myr), rank among the
youngest Earth-sized planets currently known, comparable to systems
such as HD~63433d \citep[1.1\,$R_\oplus$,
$414$$\pm$$23$\,Myr;][]{2024AJ....167...54C} and TOI-1807~b
\citep[1.3\,$R_\oplus$, 180$\pm$40\,Myr;][]{2021AJ....162...54H}.
Kepler-1629~b, with a size just two-thirds that of Earth
(0.67$\pm$0.05\,$R_\oplus$), is marginally older, also based on
rotation (\kepsixteentwonine\,Myr).  The lithium age from a
reconnaissance TRES spectrum confirms the point.

@@

Most of the stars observed by Kepler are in the thin disk.  This can
be verified by following \citet{Gaia_2018}, and denoting stars with 2D
tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
For all stars in the KIC, this yields 95{,}694 thin and 54{,}039 thick
disk stars respectively.  The thin disk stars have a threefold larger
detected rotation period fraction: 35{,}675 rotators are in the thin
disk, while 7{,}312 are in the thick disk.  Further imposing the
gyrochronology quality flags discussed in Section~\ref{subsec:flags}
yields 17{,}755 thin-disk stars for which gyrochronology is nominally
applicable, and 2{,}462 thick-disk stars.

@@

% compare against Berger+20, note their sentence
% "Encouragingly, the distribution also qualita- tively matches the red giant asteroseismology-derived age distributions in Silva Aguirre et al. (2018) and Pinsonneault et al. (2018), as well as the rotation-based ages in Claytor et al. (2020) and the Galactic Archaeology with HERMES–Gaia ages in Buder et al. (2019)."
%
% also compare against Mor2019 and their literature assessment
% "Our findings that the thin disc SFH does not follow a sim- ple decreasing shape until the present are in good agreement with Snaith et al. (2015), and Haywood et al. (2016, 2018) who found, using data with metallicities and assuming a fixed IMF, the existence of an SFR quenching followed by a reactivation. Kroupa (2002a), using stellar kinematics, found the SFH to behave similarly. The relative maximum of the SFR that we find at 2–3 Gyr ago is compatible with the results of Vergely et al. (2002) and Cignoni et al. (2006) that, using Hipparcos data in a sphere of 80 pc around the Sun and assuming a fixed IMF, found maximum peaks at 1.75–2 Gyr ago and 2–3 Gyr ago, respec- tively. Recently, Bernard (2018), in a tentative work using TGAS data, pointed towards the existence of a relative maximum also located 2–3 Gyr ago."

@@

{\it Occurrence Rates}:
The age-dependent trends predicted for exoplanet populations, such as
the Kelvin-Helmholtz cooling of mini-Neptunes \citep{Gupta_2019},
time-dependent carving of the photoevaporation desert
\citep{Owen2018}, and the time evolution of the radius valley
\citep{Rogers_2021} can be explored using our data.  We defer this
analysis to a separate publication.
Some care is required since beyond age,
exoplanet demographics also depend on stellar metallicity and mass
\citep[e.g.][]{Petigura_2018,Miyazaki2023}.

@@

We began this study with two questions: how wrong is the assumption
of a uniform age distribution for stars in the galactic thin disk?
And why are only $\approx$50 sub-gigayear transiting planets known,
rather than the $\approx$500 that would be expected under the
assumption of a uniform star formation history for the
$\approx$5{,}000 known planets?

@@

Our analysis blindly recovered the ages of all 14 known Kepler planets
in clusters.  While lithium provided only marginal added information
for most of the sample, the lithium and rotation ages agreed in
$\gtrsim$90\% of cases for which comparison was possible.
Our resulting ages included two-sided $t_{\rm gyro}$ and $t_{\rm Li}$
for \ltonegyrhighqconfirmedtwosided\ sub-gigayear confirmed planets, and
two-sided $t_{\rm gyro}$ with one-sided $t_{\rm Li}$ for
\ltonegyrhighqconfirmedonesided\ sub-gigayear confirmed planets.  Allowing for
systems with RUWE$>$1.4 and including ``candidate'' planets
would increase the counts to
\ltonegyrmediumqconfirmedtwosided\ and
\ltonegyrmediumqconfirmedonesided\ respectively; weakening other quality flags,
such as on transit geometry, would further increase the counts.  The sizes and
periods of the most restricted set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets are
mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert ''\citep{Owen2018}, and others include Earth-sized 
planets a few hundred million years old.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item The age distributions of both the Kepler target stars and the
    known Kepler planet-hosts show a demographic cliff.  There are
    \ratioobtoybstars\ times as many ``old'' (2-3\,Gyr) stars in the
    Kepler field as ``young'' (0-1\,Gyr) stars.  The star formation
    rate today is \ratiosfr$\pm$\uncratiosfr\ times lower than it was
    three billion years ago.
    This result from rotation-based ages broadly agrees with recent reports of
    a declining star-formation rate
    from CMD fitting and white-dwarf chronology, though with
    important nuances.
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the
    Kepler field implies that we should instead expect $\approx$250.
  %
  %FIXME: 120 number
  \item Using rotation-based ages, we have derived ages for
    \nplyounggyrotwosigma\ Kepler planets younger than 1\,Gyr at 2$\sigma$,
    and \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$120-170 known sub-gigayear planets with well-constrained
    ages.  This lessens the original factor of ten discrepancy to
    a factor of at most two.
\end{enumerate}

@@

\appendix
\section{Defining An ``Isochronally Too Far From The Main Sequence'' Boundary}
The (arbitrary) method used to construct the orange line in 
the lower panel of Figure~\ref{fig:stellarprops} was as follows.
The line is	defined as
the maximum of two numerically-determined functions, $f_0(T_{\rm eff})$ and $f_1(T_{\rm eff})$.
We defined $f_0$ by fitting an
$N^{\rm th}$-order polynomial to the stars with rotation periods for which $t_{\rm
	B20,iso}/(+\sigma_{t,{\rm B20,iso}})\approx 3$
and $T_{\rm eff}\in[3700,5900]$\,K, where
$t_{\rm B20,iso}$ was the isochronal age reported by
\citet{Berger_2020a_catalog}.  We let the order of the fit vary from
$N$=1 to 10, minimized the Bayesian Information Criterion,
and found $N$=6.  
By eye, this yielded a plausible locus for $T_{\rm eff}$$\lesssim$5300\,K.
However for F and G dwarfs, the resulting locus
allowed for stars that were ``too evolved''; the density of KOIs 
was anomalously low in this region of the $\log g$ vs.\ $T$$_{\rm eff}$ plane.
We therefore visually selected KOIs that appeared
to be near the main sequence -- i.e.~ most of the yellow
points in Figure~\ref{fig:stellarprops} -- and used them to fit a separate
polynomial $f_1$ through a similar BIC-minimization procedure,
which yielded $N$=3.
This polynomial, $f_1$, is shown in very faint opacity in Figure~\ref{fig:stellarprops}.
The portion of the orange locus from $\approx$5300--6200\,K
is set as $f_1 + c$, for $c$ a constant offset that we defined to be
0.1\,dex above the ``KOI main sequence''.
The exact break-point in temperature is automatically set by ${\rm max}(f_0, f_1+c)$.
We do report rotation-based ages for stars
above this orange locus;
they are simply flagged as not being near the main sequence.

\section{Cases With ``Inconsistent'' Rotation and Lithium-Based Ages}
\label{app:inconsistent}

@@

\paragraph{Kepler-1}---TrES-2/Kepler-1 \citep{2006ApJ...651L..61O} is
a $\approx$2.5\,day near-grazing hot Jupiter oribiting a G0 V primary
which has been studied in detail by multiple investigators
\citep[e.g.][]{2007ApJ...664.1190S,2008ApJ...682.1283W,2011ApJ...733...36K,2011MNRAS.417.2166S}.
Our lithium-based age of this system (\trestwotli\,Myr), derived from
EW$_{\rm Li}^{\ast}$=81$\pm$3\,m\AA, qualitatively agrees with the
previously noted lithium abundance \citep{2007ApJ...664.1190S}.
However, we detect no photometric rotation signal, and the
spectroscopic $v\sin i$ is low.  The lithium age is strongly
asymmetric because of the scatter in EW$_{\rm Li}$ vs.~$T_{\rm eff}$
over ages $t$$\gtrsim$1\,Gyr for early G dwarfs.  We calculate a
$2\sigma$ upper limit on $t_{\rm Li}$ of 2.8\,Gyr, and a $3\sigma$
upper limit of 6.1\,Gyr.  The non-detection of rotation, while mildly
surprising, is not shocking if the system's true age is in this long
tail.

\paragraph{Kepler-101}---This $\approx$5500\,K star has EW$_{\rm
Li}^{\ast}$=100$\pm$5\,m\AA, nominally implying $t_{\rm
Li}$$\approx$200-800\,Myr, but no rotation detection.  Our automated
flags noted that this star has a low surface gravity, high luminosity,
and is far from the main sequence.  \citet{2014A&A...572A...2B}
concur: this star is a sub-giant slightly more massive than the Sun;
no rotation detection is expected; the lithium likely simply survived
over the star's main sequence lifetime.

@@

\paragraph{Kepler-786}---This K3 V star has a surprisingly high
lithium content.  The spectrum is single-lined, with
\texttt{SpecMatch-Synth} derived parameters of $T_{\rm
eff}$=4769$\pm$100\,K, $\log g$=4.5$\pm$0.1, [Fe/H]=0.11$\pm$0.06, and
$v\sin i$=2.4$\pm$0.6\kms.  Yet with EW$_{\rm
Li}^{\ast}$=83$\pm$6\,m\AA, the lithium age of \kepseveneightsix\,Myr
would predict an obvious rotation signal.  None is present in the
Kepler light curve, consistent with the low $v\sin i$.  Out of all
``apparently discrepant'' lithium and rotation measurements discussed
in this appendix, this is the only one that seems to remain discrepant
after scrutiny.  The  \ion{Ca}{2} doublet is in emission, with
$R'_{\rm HK}=-4.7\pm0.5$, which suggests an age at least as old as the
Hyades \citep{Mamajek_2008}.  The Balmer lines are in absorption, and
display no obvious signatures of youth.

@@

%\section{TODO's}
%
%\begin{enumerate}
%%  \item Compare Mathur2023 ages against my ages, and (more interestingly) against the
%%    cluster scale using NGC6811, Theia520, Melange3, and CepHer stars.
%%    Could compare against asteroseis too.
%%    %
%%  \item Intro needs better discussion of Mathur2023 and Lu2021
%    %
%  \item Consider moving the bit list description to its own appendix.
%    %
%  \item Rerun gyro calculation with denser grid, fixing quantization
%    and extending upper limit to 5gyr.  (Will use updated Teff
%    uncertainties)
%\end{enumerate}
